Characterizing Stellar Streams by ML

### 2023-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

\section{Introduction} \label{sec:intro}
    Stellar streams are the disrupted remnants of globular clusters and satellite galaxies, and their census has grown extensively in the Milky Way thanks to all-sky astrometric missions like {\it Gaia} and low-surface brightness photometric surveys (\JN{cite}). The formation of streams can be traced back to a progenitor cluster (i.e., a satellite galaxy or globular cluster), whose stars are gradually lost to the tidal field of the host galaxy and extend along a series of similar orbits producing tidal tails which can sometimes span many tens of degrees across the sky. This orbital proximity of stars belonging to a stream makes them sensitive detectors of the mass distribution of galaxies, sourced by both the baryonic and dark matter components across a large range of mass-scales.

    The average properties of a stream (e.g., the mean position and velocity track of the tidal tails) can be used to reconstruct the global dark matter distribution of the stellar halo (e.g., \JN{cite}), while local variations in the stream track and surface density encode information about the stream's encounters with baryonic structures such as (e.g.) the stellar bar (\JN{cite}), and non-luminous structures like dark matter subhalos (\JN{cite}).

@@

@@

        The paper is organized as follows.
        In \autoref{sec:method} we introduce our method for characterizing stream tracks and computing stellar membership.
        We discuss the data in \autoref{sub:framing_the_data} and build a probabilistic framework to model the stream and background in \autoref{sub:likelihood_setup}-\autoref{sub:priors}.
        In \autoref{sec:results} we apply our method,
        % TODO: rephrase the mock data
        first to simulated data (\autoref{sub:mock_data}) to demonstrate the model with a known ground truth dataset,
        then to the stellar streams \stream{GD-1} (\autoref{sub:gd1}) and \stream{Pal\,5} (\autoref{sub:pal5}).
        We compare our approach to characterizing stellar streams to other methods in \autoref{sec:comparison}.
        We conclude in \autoref{sec:conclusions}, discussing the results of this work and future directions.

@@

    We now introduce our method for characterizing stellar stream tracks and computing stream membership likelihoods from N-D phase-space measurements.
    % TODO: rephrase following sentence
    We develop a flexible Bayesian model framework, that allows arbitrary numbers of phase-space dimensions to be incorporated in the model and permits missing phase-space observations. 
    Below, we offer a concise overview of our approach, followed by a more in-depth technical explanation in subsequent sections.
    We start, in \autoref{sub:framing_the_data}, by transforming the astrometric data into
    a stream-oriented on-sky reference frame ($\phi_1, \phi_2, \parallax, \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for each stream. In this frame, 
    the stream track lies along $\phi_1$, and the stream density may be described by distributions in the other parameters as injective functions of $\phi_1$.
    % TODO: ⬆ I think a few more sentences explaining why this is important, feature variations, etc?
    % However, previously we discussed how stream will vary along the track, with changes in the width and possible gaps.
    % Also observational effects, e.g the distance gradient changes the on-sky width (should this be mentioned earlier?)
    Likewise, the background of the Galactic field may be described with $\phi_1$-conditioned distributions.
    % TODO: ⬆ I think this should be expanded.
    Even non-astrometric coordinates, e.g. photometry and metallicity, are modelled along the stream and background as functions of $\phi_1$.

@@

        As a stream progenitor orbits the Galactic center of mass (GCOM), its orbital path
        describes an ellipse in the orbital plane. The associated arms of the stream generally
        align closely with its orbital plane.
        From a Galactocentric reference frame, an orbital plane assumes the form of a great
        circle in the observer's celestial "sky," thereby allowing the potential realignment of
        longitudinal ($\ell$) and latitudinal ($b$) coordinates to a coordinate system aligned
        with the stream ($\cmp{GCOM}{\phi_1}, \phi_2^{(GCOM)}$). Here, the progenitor is
        established at the origin, with the stream progressing along
        $\cmp{GCOM}{\phi_1} (\phi_2^{(GCOM)} \approx 0)$. $\cmp{GCOM}{\phi_1}$ is clearly a good
        affine parameter along which to construct models and analyze the stream's characteristics.

        However, data observations are not taken from the GCOM, but from  $\sim 8$ kpc distant
        \citep{GRAVITY2018, Leung2022} in a local frame, typically from vantage points such as
        Earth or its solar Lagrange points, and subsequently represented in the International
        Celestial Reference System (ICRS) \citep{ICRS1997}.
        Restricted as we are to a very local volume, there is no means to observe systems from
        the more ``intrinsic" reference frames. In our observational frames there is no inherent
        assurance of effecting a rotation that would similarly align the stream within observer
        coordinates such that $\phi_2(\phi_1) \approx 0$.
        Thankfully, most streams are adequately distant from both the Galactic center and the
        observer, or exhibit orientations such that a rotation a) can be established or
        b) holds valid.

        Within a Galactocentric framework and expressed in spherical coordinates, streams
        effectively approximate linear formations on a great circle, wherein a phase angle
        traverses the entire stream length.

        % Data observations are taken in an (approximately) heliocentric reference frame, e.g. from the Earth or a Lagrange point, and is reported in an ICRS
        % \citep{ICRS1997} frame. Restricted as we are to a very local volume, there
        % is no means to observe systems from more ``intrinsic" reference frames, namely from the Galactic center of mass about which orbit satellites such as stellar streams.

        % As the progenitor orbits the Galactic COM (center of mass), its
        % orbital path describes an elliptical path in an orbital plane. The
        % progenitor's tidal tails (stream) will likewise lie near the progenitor's
        % orbital plane. Observing from the Galactic center, an orbital plane lies on a
        % great circle in the observer's `sky' and it is possible to re-orient the
        % longitude ($\ell$) and latitude ($b$) to a stream-oriented coordinate system
        % ($\phi_1, \phi_2$) with the progenitor at the origin and the stream along
        % $\phi_1 (\phi_2 \approx 0)$. However, our observations are not from the
        % Galactic center, but approximately 8 kpc distant \citep[for recent
        % measurements, see][]{GRAVITY2018, Leung2022}. For an arbitrary stream lying
        % (approximately) along a galactocentric ellipse there is no guaranteed rotation to similarly align the stream in the observer coordinates so that the stream
        % lies along $\phi_2(\phi_1) \approx 0$. Thankfully, many streams are sufficiently
        % distant from both the Galaxy and observer or oriented such that a rotation a)
        % exists, or b) applies to significant lengths of the stream before deviations
        % become problematic.

        % In a Galactocentric frame and in spherical coordinates  streams are approximately lines on a great circle, with a phase-angle sweeping the length of the stream. This, clearly, is a good affine parameter along which to model and study the stream.

        {\color{red} Assume we know a stream track}

        Instead we will approach this in the standard way: by parameterizing the stream along its the longitude -- $w_{\cdot, j\equiv0} \equiv \phi_1$ -- in a stream-oriented reference frame.
        $\phi_1$ becomes our conditioning variable.

        \begin{quotation}
            The stream is projected from its initial reference frame, e.g. ICRS
            \citep{ICRS1997}, into a rotated on-sky frame, with longitude $\phi_1$ and
            latitude $\phi_2$, such that the stream extends along $\phi_1$. In this frame
            the stream may be fit by a low-order polynomial as a function of $\phi_1$
            \citep[e.g.][]{Bonaca2020}. Similar, if more sophisticated, methods swap tools
            like univariate splines \citep[e.g.][]{Erkal2017, Bonaca2018} for the
            low-order polynomials. Using a coordinate dimension, here $\phi_1$, as the
            independent variable works for many streams, as $\phi_1$ sweeps monotonically
            along the stream. For example \stream{GD-1} is straight for over 80 degrees
            \citep{Webb2019, Price-Whelan2018}.
        \end{quotation}

        There are known to be issues with this...

        \begin{quotation}
            However, this is not true in general: a
            stream's path can become kinked by subhalo interactions, or curved on the
            celestial sphere in an observer's reference frame, or like the Magellanic
            stream wrap multiple times around the Milky Way \citep{Wannier1972}. When a
            stream track is not a function, methods fitting $\phi_2(\phi_1)$ fail.
        \end{quotation}

        We leave extending the model framework to phase-wrapped, doubled-backed paths, etc to a future work.

@@

        \begin{figure}
            \centering
            \includegraphics[scale=0.5]{static/model/DAG_StreamDensity.pdf}
            \caption{Probabilistic Graphical Model of the Mixture Density Network (MDN)}
            \todo[inline]{todo}
            \begin{enumerate}
                \item Do this with \url{https://github.com/daft-dev/daft}
            \end{enumerate}
            \label{fig:DAG}
        \end{figure}

        Mixture models (MMs) are a statistical method to represent a population as a set of a sub-populations specified by probability distributions. These distributions are characterized by their parameters and the mixture by a corresponding set of mixture coefficients. Scalar mixture models may be extended to model completely general conditional distributions by making the distribution parameters and mixture coefficients functions of a conditional input \citep{McLachlan1989MixtureM}.
        Mixture Density Networks (MDNs) are a machine learning method to implement these general MM, taking the parameters and coefficients to be the outputs of feed-forward neural networks and as such, general continuous functions \citep{Bishop1994}. Functional parameters allows the model to capture variation over those parameters that is not possible with scalar-valued models. Moreover, by using neural networks the inferred functional forms are ``model-free" and driven entirely by the data.

        For the MDN we take $x_{\cdot, 0}$, which is the first feature column of the data $\mbf{X}$, as the `independent' coordinate over which the MDN parameters are conditioned. In the context of streams $x_{\cdot, 0} \equiv \phi_1$ is the longitude coordinate in a reference frame rotated such that the stream is aligned along $\phi_1$. The parameters of the stream, the background field, and the mixing coefficients of the two are now general data-driven continuous functions along $\phi_1$.
        {\color{cyan} \\ connecting to intro...

            The hypertube idea is connected to orbits.
            One might think then that a variable like the phase-angle or the arc-length, or any variable that directly sweeps along the stream would be best to use as the conditioning variable.
            This is not the case for a number of reasons (see \cite{StarkmanEtAl2023} for some discussion on this topic), not least of which is that we do not have any of these quantities. We do not know the stream track, it's what we want to know.
        }

        The PDF of a general MDN is
        \begin{equation} \label{eq:general_mixture_network}
            \!\!\! \pdf(\nth{\mbs{x}} | \mbs{\theta}(\phi_1))
            \!=\! \sum_{\mrm{q} \in \mbs{I_Q}} \! \qth{f}(\phi_1) \cmp{q}{\pdf}(\nth{\mbs{x}}|\qth{\mbs{\theta}}(\phi_1))
            % \prior(\nth{\mbs{x}}, \phi_1)
        \end{equation}
        where $\pdf_q$ is the PDF of the $q$-th model and $f_q$ its mixing coefficient for all $q$ models in the set of models indexed by $\mbs{I_Q}$. $\mbs{\theta}$ is the $\phi_1$-conditional output of the feed-forward neural network, and $\nth{\mbs{x}}$ the data on the $n$-th star in $\mbs{X}$, not including $\phi_1$. For convenience we drop the implied subscript $n$ in $\phi_1$. A full detailing of notation and symbols is included in \autoref{app:notation}. For convenience we drop the explicit notation of $\mbs{\theta}$'s dependence on $\phi_1$ and the index set $\mbs{I}_X$ in summations over its elements $x$.

@@

    \subsection{Determination of the Mixture Weight} \label{sub:mixture_weight}
        The mixture weight $f_q(\phi_1)$ as used in \autoref{eq:general_model} is characterized as the fraction of stars belonging to the mixture component $q$ over an increment $[\phi_1, \phi_1 + \dif{\phi_1}]$. Thus, the mixture weight encodes information about the linear density of the stream, which has been shown to depend on substructure in the galaxy (CITE-x). The dependence of the mixture weight on the linear density is captured by the expression
        \begin{equation}\label{eq:fraction_param}
            f_q\left(w_{i,0}\right) = \frac{N_q\left(\phi_1\right)}{N_{\rm tot}\left(\phi_1\right)},
        \end{equation}
        where $N_q\left(\phi_1\right)$ is the number of stars in component $q$ within an interval $[\phi_1, \phi_1 + \dif{\phi_1}]$, and 
        $N_{\rm tot}$ is the total number of stars in the same band (including those in component $q$). The linear  density for component $q$ is then entirely specified by the function $N_q(\phi_1)$.

@@

        In particular, for a given field the distribution of $\phi_1$ coordinates is specified by the set $\{\phi_1\}_f$. A normalizng flow is trained to this dataset, and will satisfy $\int d\phi_1 P_{\phi_f}(\phi_1) = 1$ by construction. In order to re-weight the normalizing flow to obtain $N_{\rm tot}(\phi_1)$, we simply rescale $P_{\phi}(\phi_1)$ by $N_{\rm tot,0}$, which is the total number of stars in the field considered. We can then obtain the linear stream density with
        \begin{equation}
            N_q\left(\phi_1\right) = f_q N_{\rm tot} \left(\phi_1\right)  = f_q \left[N_{\rm tot,0} P_{\phi_1}\left(\phi_1\right) \right] .
        \end{equation}

@@

            The astrometric track is modeled with a neural network, encompassing each phase space dimension, parameterized as a function of $\phi_1$. The neural network output for a single stream component is the mean location of the stream track in each dimension, $\astroM{\mbs{\mu}}(\phi_1)$, and its associated covariance matrix $\astroM{\mbs{\Sigma}}(\phi_1)$. The probability for the data is then
            \begin{equation}
                \cmp{S,w}{\pdf}(\nth{\mbs{w}} | \mbs{\theta}) \! = \mcal{N}_F \big(\nth{\mbs{w}} | \astroM{\mbs{\mu}}\!(\phi_1), {\cmp{w}{\nth{\mbs{\Sigma}}}}^2 \! + \!  \astroM{\mbs{\Sigma}}^2\!\left(\phi_1\right) \!\big),
            \end{equation}
            where $\astroM{\mbs{\theta}}$ is the set of neural network parameters that define $\astroM{\mbs{\mu}}$ and $\astroM{\mbs{\Sigma}}$; and $\cmp{w}{\nth{\mbs{\Sigma}}}$ the data uncertainties.
            In the simplest case, we take $\astroM{\mbs{\Sigma}}$ to be diagonal, consisting of both intrinsic dispersion for each dimension of $\mbs{w}$ and the data uncertainties added in quadrature. The data uncertainties $\cmp{w}{\nth{\mbs{\Sigma}}}$ are drawn from $\mbs{\delta}\mbf{W}$, the dataset of astrometric uncertainties (see \autoref{app:notation}).

@@

            In the presence of missing phase-space coordinate $f_k$ we reduce the dimensionality of the Gaussian to that of the data-vector. Namely, for $\mbs{I}_{\tilde{F}} = \mbs{I}_{F} \backslash \{f_k\}$
            \begin{align}
                \nth{\tilde{\mbf{w}}} &\mapsto \nth{\tilde{\mbf{w}}} = \{ w_{n,f} | f\in \mbs{I}_{\tilde{F}}\}, \nonumber\\
                \astroM{\mbs{\mu}} &\mapsto \astroM{\tilde{\mbs{\mu}}} = \{ \mu_{f} | f\in \mbs{I}_{\tilde{F}}\}, \nonumber\\
                \astroM{\mbs{\Sigma}} &\mapsto \astroM{\tilde{\mbs{\Sigma}}} = \{ \Sigma_{i,j} | i,j \in \mbs{I}_{\tilde{F}} \}, & \nonumber
                \intertext{such that,}
                \Spdf(\nth{\mbs{w}} & | \astroM{\mbs{\theta}}(\phi_1))
                    \mapsto \Spdf(\nth{\tilde{\mbs{w}}} | \astroM{\tilde{\mbs{\theta}}} (\phi_1)) = \mcal{N}_{\tilde{F}}.
            \end{align}

@@

            For the $n$-th star, the photometric model is:
            \begin{multline} \label{eq:photometric_probability_point}
                \cmp{S,m}{P}(\mbs{m_i}, \gamma | \mbs{\theta}(\phi_1)) 
                \\ \sim \mcal{N}(\nth{\mbs{m}} | \{\mcal{I(\gamma)} + \mu(\phi_1), \mbs{\Sigma}_{\mcal{I}}^2(\gamma) + \mathds{1} \sigma_\mu^2(\phi_1) + {\cmp{m}{\nth{\mbs{\Sigma}}}}^2 \}) \\ \times \prior(\gamma, \nth{\mbs{m}}, \phi_1),
            \end{multline}

@@

            We are interested in the total probability (at some $\phi_1$) and so integrate over $\gamma$ to find:
            \begin{small}
            \begin{align} \label{eq:photometric_probability}
                &\cmp{S,m}{P}(\nth{\mbs{m}} | \mbs{\theta}(\phi_1)) = \\
                &\ \int\limits_{\gamma=0}^{1} \mcal{N}(\nth{\mbs{m}} | \{\mcal{I(\gamma)} + \mu(\phi_1), \mbs{\Sigma}_{\mcal{I}}^2(\gamma) + \mathds{1}\sigma_\mu^2(\phi_1) + {\cmp{m}{\nth{\mbs{\Sigma}}}}^2 \}) \nonumber
                \\
                &\qquad \times \prior(\gamma, \nth{\mbs{m}}, \phi_1) \rm{d}\gamma. \nonumber
            \end{align}
            \end{small}
            Then, for a given $\phi_1$ we can obtain a distribution over the distance modulus using 
            \begin{align}
                \mu_{\rm sample}\left(\phi_1\right) \sim \mathcal{N}\left(\mu(\phi_1), \sigma^2_\mu(\phi_1) \right) & \quad\unit{mag}.
            \end{align}
            For each sample, we can convert the distance-modulus distribution (and error) to the distance track using 
            \begin{align}
                \rm{dist}\left(\phi_1\right) &= 10^{\frac{1}{5}\mu_{\rm sample}(\phi_1)+1} & \quad\unit{pc}, \\
                \delta \rm{dist}(\phi_1) &\simeq \frac{\ln{10}}{5} \rm{dist}(\phi_1) \ \sigma_\mu(\phi_1) & \quad\unit{pc}.
            \end{align}

@@

    % \vspace{10pt}

    % \subsection[Astrometric Background]{Astrometric Background: pieces of $\Bcmp{P}(\mbs{w})$} \label{sub:background_astrometric}

    %     When looking for a stellar stream, the highly complex Galaxy is the background. There are numerous approaches to modeling a complex structure, from a mix of very simple models to a single highly flexible model, like a basis function expansion or ML network.
    %     We aim towards the mix of simple models, but use more flexible models as needed. In this section we describe the various models used when modeling the background, in different coordinates of the phase space.

        % \subsubsection{Flat Background} \label{ssub:flat_background}

>>>>

\section{Introduction} \label{sec:intro}
    Stellar streams are the disrupted remnants of globular clusters and satellite galaxies, and their census has grown extensively in the Milky Way thanks to all-sky astrometric missions like {\it Gaia} and low-surface brightness photometric surveys (\JN{cite}). The formation of streams can be traced back to a progenitor cluster (i.e., a satellite galaxy or globular cluster), whose stars are gradually lost to the tidal field of the host galaxy and extend along a series of similar orbits producing tidal tails, which can sometimes span many tens of degrees across the sky. This orbital proximity of stars belonging to a stream makes them sensitive tracers of the mass distribution of galaxies, sourced by both the baryonic and dark matter components.

    The average properties of a stream (e.g., the mean position and velocity track of the tidal tails) can be used to reconstruct the global dark matter distribution of the stellar halo (e.g., \JN{cite}), while local variations in the stream track and surface density encode information about the stream's encounters with baryonic structures such as (e.g.) the stellar bar (\JN{cite}), and non-luminous structures like dark matter subhalos (\JN{cite}). Importantly, these galactic components span orders of magnitude in mass, highlighting the importance of characterizing a population of stellar streams both in morphology and kinematics.

@@

    While streams are well described in (e.g.) action angle coordinates or $N-$body simulations, these methods relying on prior assumptions about the galaxy and its underlying potential. Therefore, it is important to devise data-driven methods to characterize stellar streams without appealing to strong prior assumptions about the galaxy's structure.

    There are a few previous works which have modeled stellar streams in this context. \citet{2022MNRAS.514.1757P} developed a data-driven spline-based model for modeling the photometric qualities of previously identified streams. Their method works by fitting the stream's stellar surface density with a series of splines, capturing both surface brightness variations along the stream and changes in the position of the stream track and width. Their method also utilize a simultaneous ischorone fit to the stream, from which a distance modulus can be derived by assuming that the entire stream originates from the same stellar population. While this method is very successful at characterizing the photometric properties of a stream, it does not consider the kinematic dimensions which have become increasing well-measured with successive data releases from {\it Gaia}, as well as targeted observations. From a population of RR Lyrae stars, \citet{2019AJ....158..223P} utilized a mixture model to fit the globular cluster stream Pal-5 in the space of proper motions and heliocentric distances. The mixture modeling approach enables a simultaneous fit to both the stream and background density, enabling better separation of a coherent stream from an otherwise noisy background distribution. However, the track of the stream in on-sky coordinates is not fit. The method identified 27 RR Lyrae stars consistent with being members of the stream. While useful for determining the average properties of the stream (e.g., its mean distance track), a limited population of RR Lyrae stars are not sufficient to capture the fine-grained properties of streams expected to inform constraints on (e.g.) a population of perturbers in the galaxy. \JN{are there other methods?}

    Other methods for extracting stellar streams from the data include matched-filter approaches, where over-densities in color-magnitude space are inspected as possible clusters undergoing tidal stripping \JN{cite the many key papers using this method}. While the matched-filter technique has been extremely successful at stream discovery, it is not equipped to characterize streams at the level of detailed necessary to inform precise constraints on the potential of the galaxy and its substructure.

    In order to address the need for a flexible model of stellar streams both in photometric and kinematic spaces, in this work we develop a new method for characterizing streams and quantifying the membership probability of possible stream stars. Our approach provides several key advantage over previous stream modeling efforts, specifically in its ability to jointly model the kinematic and photometric properties of streams simultaneously. Furthermore, our method is well-suited to fitting streams with only partial or incomplete phase-space or photometric observations, enabling a full characterization of streams insofar as the available data allows.

@@

        The paper is organized as follows.
        In \autoref{sec:method} we introduce our method for characterizing stream tracks and computing stellar membership.
        We discuss the data in \autoref{sub:framing_the_data} and build a probabilistic framework to model the stream and background in \autoref{sub:likelihood_setup}-\autoref{sub:priors}.
        In \autoref{sec:results} we apply our method,
        % TODO: rephrase the mock data
        first to simulated data (\autoref{sub:mock_data}) to demonstrate the model with a known ground truth dataset,
        then to the stellar streams \stream{GD-1} (\autoref{sec:results_gd1}) and \stream{Pal\,5} (\autoref{sub:pal5}).
        We compare our approach to characterizing stellar streams to other methods in \autoref{sec:comparison}.
        We conclude in \autoref{sec:conclusions}, discussing the results of this work and future directions.

@@

    We now introduce our method for characterizing stellar stream tracks and computing stream membership likelihoods from N-D phase-space measurements.
    % TODO: rephrase following sentence
    We develop a flexible Bayesian model framework, that allows arbitrary numbers of phase-space dimensions to be incorporated in the model and permits missing phase-space observations. 
    Below, we offer a concise overview of our approach, followed by a more in-depth technical explanation in subsequent sections.
    We start, in \autoref{sub:framing_the_data}, by transforming the astrometric data into
    a stream-oriented on-sky reference frame ($\phi_1, \phi_2, \parallax, \mu_{\phi_1}^*, \mu_{\phi_2}, v_r)$, for each stream. In this frame, 
    the stream track lies along $\phi_1$, and the stream density may be described by distributions in the other parameters as injective functions of $\phi_1$.
    % TODO: ⬆ I think a few more sentences explaining why this is important, feature variations, etc?
    % However, previously we discussed how stream will vary along the track, with changes in the width and possible gaps.
    % Also observational effects, e.g the distance gradient changes the on-sky width (should this be mentioned earlier?)
    Likewise, the background of the Galactic field may be described with $\phi_1$-conditioned distributions.
    % TODO: ⬆ I think this should be expanded.
    Non-astrometric coordinates, e.g. photometry and metallicity, are also modelled along the stream and background as functions of $\phi_1$.

@@

        As a stream progenitor orbits the Galactic center of mass (GCOM), its orbital path
        describes an ellipse in the orbital plane. The associated arms of the stream generally
        align closely with its orbital plane.
        From a Galactocentric reference frame, an orbital plane assumes the form of a great
        circle in the observer's celestial ``sky," thereby allowing the potential realignment of
        longitudinal ($\ell$) and latitudinal ($b$) coordinates to a coordinate system aligned
        with the stream ($\cmp{GCOM}{\phi_1}, \phi_2^{(GCOM)}$). Here, the progenitor is
        established at the origin, with the stream progressing along
        $\cmp{GCOM}{\phi_1} (\phi_2^{(GCOM)} \approx 0)$. $\cmp{GCOM}{\phi_1}$ is clearly a good
        affine parameter along which to construct models and analyze the stream's characteristics.

        However, data observations are not taken from the GCOM, but from  $\sim 8$ kpc distant
        \citep{GRAVITY2018, Leung2022} in a local frame, typically from vantage points such as
        Earth or its solar Lagrange points, and subsequently represented in the International
        Celestial Reference System (ICRS) \citep{ICRS1997}.
        Restricted as we are to a very local volume, there is no means to observe systems from
        the more ``intrinsic" reference frames. In our observational frames there is no inherent
        assurance of effecting a rotation that would similarly align the stream within observer
        coordinates such that $\phi_2(\phi_1) \approx 0$.
        Thankfully, most streams are adequately distant from both the Galactic center and the
        observer, or exhibit orientations such that a rotation a) can be established or
        b) holds valid. For most of the best-studied globular cluster streams -- \stream{GD-1},
        \stream{Palomar 5}, \stream{Jhelum}, etc. -- this holds true. However for some streams
        of dwarf galaxies, like the Magellanic stream which wraps multiple times around the
        Milky Way \citep{Wannier1972}, no such transformation is possible. More generally,
        a stream's ``sky"-projected path may be kinked, e.g. by subhalo interactions, such
        that it is a many-to-one function in $\phi_1$.

        In this work we treat with the vast majority of streams, which have a ``sky" transformation such that the stream is an injective function in $\phi_2(\phi_1)$.
        Since we are examining known streams this transformation is known \textit{a priori}, or
        may be constructed during pre-processing \citep[see][\S2.1]{StarkmanEtAl2023}.
        We note that, with sufficient prior knowledge, many-to-one functions may be broken into injective segments and their connection constrained by a prior. Thus the methods
        developed subsequently may extented and applied even to wrapped and kinked streams. However, 
        we leave extending the model framework to a future work.

@@

        \begin{figure}
            \centering
            \includegraphics[scale=0.5]{static/model/DAG_StreamDensity.pdf}
            \caption{Probabilistic Graphical Model of the Mixture Density Network (MDN)}
            \todo[inline]{todo}
            \begin{enumerate}
                \item remake this example this with \url{https://github.com/daft-dev/daft}.
            \end{enumerate}
            \label{fig:DAG}
        \end{figure}

        Mixture models (MMs) are a statistical method to represent a population as a set of a sub-populations specified by probability distributions. These distributions are characterized by their parameters and the mixture by a corresponding set of mixture coefficients. Scalar mixture models may be extended to model completely general conditional distributions by making the distribution parameters and mixture coefficients into functions of a conditional input \citep{McLachlan1989MixtureM}.
        Mixture Density Networks (MDNs) are a machine learning method to implement these general MM, taking the parameters and coefficients to be the outputs of feed-forward neural networks and as such, general continuous functions \citep{Bishop1994}. Functional parameters allow the model to capture variation over those parameters that is not possible with scalar-valued models. Moreover, by using neural networks the inferred functional forms are ``model-free" and driven entirely by the data.

        For the MDN we take $x_{\cdot, 0}$, which is the first feature column of the data $\mbf{X}$, as the `independent' coordinate over which the MDN parameters are conditioned. In the context of streams $x_{\cdot, 0} \equiv \phi_1$ is the longitude coordinate in a reference frame rotated such that the stream is aligned along $\phi_1$. The parameters of the stream, the background field, and the mixing coefficients of the two are now general data-driven continuous functions along $\phi_1$.

        The PDF of a general MDN is
        \begin{equation} \label{eq:general_mixture_network}
            \!\!\! \pdf(\nth{\mbs{x}} | \mbs{\theta}(\phi_1))
            \!=\! \sum_{\mrm{q} \in \mbs{I_Q}} \! \qth{f}(\phi_1) \cmp{q}{\pdf}(\nth{\mbs{x}}|\qth{\mbs{\theta}}(\phi_1)),
            % \prior(\nth{\mbs{x}}, \phi_1)
        \end{equation}
        where $\cmp{q}{\pdf}$ is the PDF of the $q$-th model and $f_q$ its mixing coefficient for all $q$ models in the set of models indexed by $\mbs{I_Q}$. $\mbs{\theta}$ is the $\phi_1$-conditional output of the feed-forward neural network, and $\nth{\mbs{x}}$ the data on the $n$-th star in $\mbs{X}$, not including $\phi_1$. For convenience we drop the implied subscript $n$ in $\phi_1$, the explicit notation of $\mbs{\theta}$'s dependence on $\phi_1$, and the index set $\mbs{I}_X$ in summations over its elements $x$. A full detailing of notation and symbols is included in \autoref{app:notation}.

@@

    \vspace{10pt}
    \subsection{Determination of the Mixture Weight} \label{sub:mixture_weight}
        The mixture weight $f_q(\phi_1)$ as used in \autoref{eq:general_model} is characterized as the fraction of stars belonging to the mixture component $q$ over an increment $[\phi_1, \phi_1 + \dif{\phi_1}]$. Thus, the mixture weight encodes information about the linear density of the stream, which has been shown to depend on substructure in the galaxy (CITE-x). The dependence of the mixture weight on the linear density is captured by the expression
        \begin{equation}\label{eq:fraction_param}
            f_q\left(w_{i,0}\right) = \frac{N_q\left(\phi_1\right)}{N_{\rm tot}\left(\phi_1\right)},
        \end{equation}
        where $N_q\left(\phi_1\right)$ is the number of stars in component $q$ within an interval $[\phi_1, \phi_1 + \dif{\phi_1}]$, and 
        $N_{\rm tot}$ is the total number of stars in the same band (including those in component $q$). The linear  density for component $q$ is then entirely specified by the function $N_q(\phi_1)$.

@@

        In particular, for a given field the distribution of $\phi_1$ coordinates is specified by the set $\{\phi_1\}_f$. A normalizng flow is trained to this dataset, and will satisfy $\int d\phi_1 P_{\phi_f}(\phi_1) = 1$ by construction. In order to re-weight the normalizing flow to obtain $N_{\rm tot}(\phi_1)$, we simply rescale $P_{\phi}(\phi_1)$ by $N_{\rm tot,0}$, which is the total number of stars in the field considered. We can then obtain the linear stream density with
        \begin{equation}
            N_q\left(\phi_1\right) = f_q N_{\rm tot} \left(\phi_1\right)  = f_q N_{\rm tot,0} P_{\phi_1}\left(\phi_1\right) .
        \end{equation}

@@

            The astrometric track is modeled with a neural network, encompassing each phase space dimension, parameterized as a function of $\phi_1$. The neural network output for a single stream component is the mean location of the stream track in each dimension, $\astroM{\mbs{\mu}}(\phi_1)$, and its associated covariance matrix $\astroM{\mbs{\Sigma}}(\phi_1)$. The probability for the data is then
            \begin{equation}
                \cmp{S,w}{\pdf}(\nth{\mbs{w}} | \mbs{\theta}) \! = \mcal{N}_F \big(\nth{\mbs{w}} | \astroM{\mbs{\mu}}\!(\phi_1), {\cmp{w}{\nth{\mbs{\Sigma}}}}^2 \! + \!  \astroM{\mbs{\Sigma}}^2\!\left(\phi_1\right) \!\big),
            \end{equation}
            where $\astroM{\mbs{\theta}}$ is the set of neural network parameters that define $\astroM{\mbs{\mu}}$ and $\astroM{\mbs{\Sigma}}$; and $\cmp{w}{\nth{\mbs{\Sigma}}}$ the data uncertainties.
            In the simplest case, we take $\cmp{w}{\nth{\mbs{\Sigma}}}, \astroM{\mbs{\Sigma}}$ to be diagonal, consisting of both intrinsic dispersion for each dimension of $\mbs{w}$ and the data uncertainties added in quadrature. The data uncertainties $\cmp{w}{\nth{\mbs{\Sigma}}}$ are drawn from $\mbs{\delta}\mbf{W}$, the dataset of astrometric uncertainties (see \autoref{app:notation}).

@@

            In the presence of missing phase-space coordinate $f_k$ we reduce the dimensionality of the Gaussian to that of the data-vector. Namely, for $\mbs{I}_{\tilde{F}} \equiv \mbs{I}_{F} \backslash \{f_k\}$
            \begin{align}
                \nth{\tilde{\mbf{w}}} &\mapsto \nth{\tilde{\mbf{w}}} = \{ w_{n,f} | f\in \mbs{I}_{\tilde{F}}\}, \nonumber\\
                \astroM{\mbs{\mu}} &\mapsto \astroM{\tilde{\mbs{\mu}}} = \{ \mu_{f} | f\in \mbs{I}_{\tilde{F}}\}, \nonumber\\
                \astroM{\mbs{\Sigma}} &\mapsto \astroM{\tilde{\mbs{\Sigma}}} = \{ \Sigma_{i,j} | i,j \in \mbs{I}_{\tilde{F}} \}, & \nonumber
                \intertext{such that,}
                \Spdf(\nth{\mbs{w}} & | \astroM{\mbs{\theta}}(\phi_1))
                    \mapsto \Spdf(\nth{\tilde{\mbs{w}}} | \astroM{\tilde{\mbs{\theta}}} (\phi_1)) = \mcal{N}_{\tilde{F}}.
            \end{align}

@@

            For the $n$-th star, the photometric model is:
            \begin{multline} \label{eq:photometric_probability_point}
                \cmp{S,m}{\pdf}(\mbs{m_i}, \gamma | \mbs{\theta}(\phi_1)) 
                \\ \sim \mcal{N}(\nth{\mbs{m}} | \{\mcal{I(\gamma)} + \mu(\phi_1), \mbs{\Sigma}_{\mcal{I}}^2(\gamma) + \mathds{1} \sigma_\mu^2(\phi_1) + {\cmp{m}{\nth{\mbs{\Sigma}}}}^2 \}) \\ \times \prior(\gamma, \nth{\mbs{m}}, \phi_1),
            \end{multline}

@@

            We are interested in the total probability (at some $\phi_1$) and so integrate over $\gamma$ to find:
            \begin{small}
            \begin{align} \label{eq:photometric_probability}
                &\cmp{S,m}{\pdf}(\nth{\mbs{m}} | \mbs{\theta}(\phi_1)) = \\
                &\ \int\limits_{\gamma=0}^{1} \mcal{N}(\nth{\mbs{m}} | \{\mcal{I(\gamma)} + \mu(\phi_1), \mbs{\Sigma}_{\mcal{I}}^2(\gamma) + \mathds{1}\sigma_\mu^2(\phi_1) + {\cmp{m}{\nth{\mbs{\Sigma}}}}^2 \}) \nonumber
                \\
                &\qquad \times \prior(\gamma, \nth{\mbs{m}}, \phi_1) \rm{d}\gamma. \nonumber
            \end{align}
            \end{small}
            Then, for a given $\phi_1$ we can obtain a distribution over the distance modulus using 
            \begin{align}
                \mu_{\rm sample}\left(\phi_1\right) \sim \mathcal{N}\left(\mu(\phi_1), \sigma^2_\mu(\phi_1) \right) & \quad\unit{mag}.
            \end{align}
            For each sample, we can convert the distance-modulus distribution (and error) to the distance track using 
            \begin{align}
                \rm{dist}\left(\phi_1\right) &= 10^{\frac{1}{5}\mu_{\rm sample}(\phi_1)+1} & \quad\unit{pc}, \\
                \delta \rm{dist}(\phi_1) &\simeq \frac{\ln{10}}{5} \rm{dist}(\phi_1) \ \sigma_\mu(\phi_1) & \quad\unit{pc}.
            \end{align}

@@
